Regression Cube Analysis of Cohort Study Data

### 2015-03-24 draft.tex
revised using Bernhards notes

%% Paper title.

@@

\maketitle
%%
Epidemiology aims to characterize health and disease conditions in defined populations (\emph{Cohorts}).
%%
Insights about risk factors allow to characterize disease-specific high-risk groups and act as important diagnostic key figures \cite{Fletcher2012}.
%%
They can also be used to give recommendations regarding a healthy lifestyle and provide information about wide spread diseases.
%%
In the standard workflow, physicians translate observations into hypotheses, which are depicted using epidemiological features and then assessed using regression analyses.

An important epidemiological tool for deriving such features are \emph{Cohort studies}, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}.
%%
To reduce any selection bias, subjects are invited at random and without a focus on a specific disease.
%%
The acquired features range from social and lifestyle factors to prior or current diseases and medications as well as medical parameters, such as blood pressure and also comprises of non-radiating medical image data. e.g. magnetic resonance imaging (MRI).
%%
Medical image data quantified using user-defined landmarks, describing for example shape, volume or diameter of a structure.
%%

To assess the statistical resilience of a hypotheses using regression analyses rarely involves more than three features due to the required subject count.
%%
Due to missing overview techniques, possibly interesting correlations lie within the data, but are not made apparent.
%%
Explorative analyses and first overview visualizations of the data set as presented by Klemm et al. \cite{Klemm2014VIS} are not custom tailored to a specific target variable and mostly highlights correlations between variables, which are known to the domain expert (e.g. correlation between body size and spine shape).
%%
We incorporate the regression analysis, which is familiar to the domain experts into a overview visualizations, which can either be used for an hypothesis-free analysis or a analysis towards a specific disease or hypotheses.
%%
This is achieved by providing template regression formulas, which are applied to all potential variable combinations.
%%
Since the notation is familiar to epidemiologists, they can rapidly include their domain knowledge into the analysis process.
%%
Difference views between regression formulas allow to assess the influences of individual variables in the process.

Our contributions are:
\begin{itemize}
	\item A overview visualization technique describing feature interactions using target features.
	\item Visualization techniques, which incorporate overview visualizations of all regression analyses at once as well as details on demand techniques for detailed investigations of feature relationships.
	\item freely adjustable regression formulas provide a simple, yet powerful way to adjust the regression analysis to specific hypotheses about the data.
	\item analysis of confounding variables by providing comparison views between different formula results
	\item The open and web based approach of the system allows for analysis of any data using the presented method.
\end{itemize}

\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow and requirements.
%%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
Epidemiology unites experts from different academic disciplines, such as physicians, statisticians and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is derive disease-specific risk factors by assessing epidemiological features using statistical methods.
%%
As described by Thew et al. \cite{Thew2009} it is divided into these different steps:
\begin{itemize}
	\item Clinicians make observations in the daily practice which are translated into hypotheses.
	\item Epidemiologists compile a list of variables depicting the hypothesis and include confounding variables.
	\item Statisticians assess the association of the derived features towards the investigated disease.
\end{itemize}
%%
Relative risks can be determined if a statistical resilient association of features towards a condition are extracted.
%%
They indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is a epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS} are used to analyzed the data using the classical sequential epidemiological workflow, which uses images largely only to communicate results, rather than providing insight.
%%
A alternative data driven approach is described as follows.

\subsubsection{Data Driven Hypothesis-Generation-based Analysis}
Klemm et al. describe an approach for a Interactive Visual Analysis approach for image-centric cohort study data, which connects to the variable listing step \cite{Klemm2014VIS}.
%%
Their methods aim to derive hypotheses through analysis of the data and observations about previously unknown feature correlations.
%%
Employing \emph{hypotheses generation} requires overview visualizations of feature correlations, which are not supported by standard statistical processors.

\subsection{Epidemiological Data} \label{sec:EpidemiologicalData}
%%
\emph{Cohort studies} are a major epidemiological tool for gathering epidemiological features.
%%
They yield a highly heterogenous and incomplete information space.
%%
Subject data are collected with the widest range possible, allowing the data set to be assessed towards different diseases.
%%
The feature space comprises information about lifestyle, somatometric variables, medical parameters, genetic data as well as medical image data derived through different modalities, such as questionnaires or medical examinations or laboratory analyses.
%%
Many features are sparse, such as follow up questions about a medication or treatment of a certain disease.
%%
Other features are exclusive, such as women-specific questions, e.g. number of born children or period status.

@@

\paragraph{Image Data.}
%%
Since the Rotterdam study, many modern cohort studies include medical image data.
%%
For ethical reasons, the imaging modalities must not include ionizing radiation.
%%
The image quality is often inferior to clinical standard, which is a tradeoff between time and cost \cite{Preim2014}.
%%
These data are hard to analyze as they require segmentation highlighting the structures of interest.
%%
This process is prone to inter and intra-observer variability when carried out manually.
%%
Automatic or semi-automatic solutions bypass this problem, but are costly and need to be custom tailored to the structure of interest \cite{Toennies2015}.
%%
Image-derived variables are either dichotomous variables indicating a medical finding by a radiologist or continuous, describing the shape of segmented structures (e.g. volume or diameter).

\paragraph{Confounding Features}
Features influencing the exposure as well as the outcome of a analysis are called confounders.
%%
The analysis model has to be adjusted by normalizing all included features towards the confounding feature.
%%
\emph{Age} is included as confounder in almost any epidemiological analysis, since most diseases (such as different cancer types) are more likely to happen with increasing age.
%%
It also influences the general body condition and therefore almost all features acquired through cohort studies.
%%
Another important confounder is \emph{Gender}.
%%
Other confounder have to be selected by epidemiologists specific to the investigated condition.

\subsection{Regression Analysis} \label{sec:RegressionAnalysis}
%%
Regression analyses are the most important statistical tool when analyzing epidemiological data.
%%
They are also the foundation of this work.
%%
A regression analysis assesses the influence of one or more (\emph{independent}) features to one target (\emph{dependent}) feature.
%%
The regression model yields a function describing the target feature by weighting the independent features.
%%
Metrics, such as the weightings itself and associated p values and describe the resulting function (the \emph{model}).
%%
R$^2$ values describe the goodness of fit; in other words how well the dependent features describe the target feature.
%%
The value is in the range [0,1], while 1 encodes a perfect fit.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany aims to aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per subject.
%%
A second cohort, \texttt{SHIP-TREND} was introduced in 2012.
%%
Data for both cohorts are examined in a in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the last acquisition (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
Tukey already stated in 1977 that data if too often only analyzed using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
We put each paper in relation to our work, derive ideas and point out differences, starting with the works closest to ours:
%%
\\\\
The work of Piringer et al. \cite{Piringer}, Chan et al. \cite{Chan}, Angelelli et al. \cite{Angelelli} and Zhang et al. \cite{Zhang2014} is closest to ours regarding different aspects.
%% HyperMoVal: Interactive Visual Validation of Regression Models for Real-Time Simulation
Akin to our approach, Piringer et al. \cite{Piringer} propose methods for visualizing regression analysis results and properties for developing car engines.
%%
Their main goal is to pairwise assess the influence of independent features towards the target.
%%
This is achieved by using a plot matrix displaying models as contours towards the target feature.
%%
They also incorporate 3D visualizations for each pairwise combination, but mainly because it is popular with the engineers in their target domains.
%%
Linked views of model deviations allow to select outlier residuals.
%%
Comparison of models is also included by mapping model values on line-style (e.g. solid for $model_a$ and dashed for $model_b$).
%%
This limits the method to comparing a few models at once.
%%
The plot matrix representation however gets complex and harder to understand with increasing number of independent features.
%%
The main difference is their focus on analyzing one complex model in detail, yielding in extensive plots, while we analyze a large amount of simpler models towards \emph{different} features.
%%
They also focus on metric features, while we process also categorical data.
%%
We share the matrix-arrangement of results depicting relationships between independent features in a model.

@@

%% TODO: Check DLPD for correct page numbers
While the presented techniques were well received with epidemiological experts, the explanatory power towards back pain was limited, yielding a analysis based on image-derived features, such as extracting and analyzing curvature, torsion and angle of the lumbar spine \cite{Klemm2015}.
%%
These measures were analyzed using a novel decision tree quality plot towards non-image data.
%%
We concluded that the model quality is not enough to characterize back pain.
\\\\
The difference of the presented approach towards our previous and other herein presented work is twofold:
\begin{enumerate}
	\item We focus on large scale analysis of a vast number of linear and logistic regression-models by assessing their quality-of-fit using descriptive metrics.
	\item The analysis is conducted towards a target feature and incorporates expert knowledge via the regression model definition rather than subdividing the underlying data.
\end{enumerate}
\com{Main Difference to our and other related work}
\section{Regression Cube Analysis of Cohort Study Data}
\com{Add Data Preprocessing Section.}
%%
The basic idea of our \emph{Regression Cube} is to provide a overview visualization of large cohort study data sets towards target variables.
%% TODO BLIND
Overview visualizations of feature relationships as presented by Klemm et al. \cite{Klemm2014VIS} are often focused on relationships between the visualized features.
%%
Correlation metrics, such as the \emph{Pearson product-moment correlation coefficient} or \emph{Cram\'{e}r's V} contingency values are incorporated to achieve this goal.
%%
In epidemiology, these relationships are also of interest, but rather towards their explanatory power towards the target feature.
%%
These target features often indicates the presence of the investigated disease.
%%
As described in Section \ref{sec:RegressionAnalysis}, regression analyses are the statistical tool of choice for analyzing these relationships.
%%
A regression model is based on expert knowledge, there is no unified rule on how to apply them to a given set of features, so they have to be applied with care.
%%
\subsection{Cube Description using Regression Formula Notation}
%%
Expert knowledge is introduced into regression analyses using the regression formulas.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the formula input influences the type of the chosen regression method as well as the \emph{independent} features describing the target.
%%
%In order to calculate all combinations of two to three independent features, we can apply  \emph{Regression Cube} aims to 
%%

Since we want to use the regression analyses with a overview visualization, we are are interested in all possible combinations of (two or more) independent features describing a target.
%%
We achieve this by introducing dynamic variables $X$, $Y$ and $Z$ into the regression notation.
%%
The method then replaces the dynamic variables with all features in the data set.
%%
In a data set with 100 features, the regression formula
\begin{equation}
Cancer \sim X + Y
\end{equation}
would yield 10.000 regression models, describing all possible combinations of two features in the data describing the target $Cancer$.
%%
The major advantage of this notation is that it comes natural to anyone familiar with regression analysis, because it uses the same notation, all operators can be used as before.
%%
This allows for a fast adaptation and in the epidemiological application domain.
%%
With simple adjustments to the formula, different results can be achieved:
%%
\begin{itemize}
	\item $Z \sim X + Y$ calculates all combinations of two features towards all possible target features.
	\item $Cancer \sim X + Y + BodyWeight$ includes the $BodyWeight$ feature into all regression models as feature.
	\item $Cancer \sim X + Y + Z$ calculates all combinations of three features towards the target.
\end{itemize}
%%
The problem with this brute-force approach lies in its complexity.
%%
The number calculated regression models increases exponentially for each dynamic variable added.
%%
If we assume a data set with 100 features with the formula $Z \sim X + Y$ we calculate 1,000,000 regression models.
%%
When each regression takes about 50~ms calculation, we have to wait roughly 14 hours for the calculation to complete.
%%
Therefore, the computational complexity needs to be reduced.
%%
%\com{Regression formulas are described using standard regression formula notation by introducing variables.}
%%
%\com{Target Group: Physicians/Epidemiologists with basic statistical background to understand regression formulas and Interaction.}

\subsection{Target-Variable-dependent Dimension Reduction}
\com{Uli, Can you please proofread this?}
%%
The vast majority of features in an epidemiological data have no or a very low relation towards the target feature.
%%
Identifying these features und excluding them from the calculation can reduce the number of dimension significantly.
%%
The \emph{Correlation based Feature Selection} (CFS) algorithm is very popular in data mining for achieving this task \cite{CFS}.
%%
It uses information entropy to select the features which have the most explanatory power towards the target feature.
%%
At the same time it tries to reduce the correlation between the independent features.
%%
When for example the \emph{body weight} has a strong explanatory power towards the target it is likely that \emph{BMI} or \emph{waist circumference} behave similar towards it.
%%
They, however, correlate strongly with each other.
%%
The CFS algorithm would then select the feature which has the largest explanatory power and discards the other dimensions.

@@

The number of features calculated by the CFS algorithm is dependent on the information entropy in the data.
%%
In our epidemiological data we observed a usual number of 10 to 20 features.
%%
For features, such as age or gender, which affect most other features (Recall "Confounder" in Section~\ref{sec:EpidemiologicalData}) the number is significantly larger (about half the features in the whole data set).
%%

@@

\subsection{Abstracting Regression Results using R$^2$}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/cube_sketch}
 \caption{
 (a) Overview visualization using a heatmap for the formula $Z \sim X + Y$, where $Z$ assumes feature $Age$.
 %%
 The R$^2$ values extracted from the regression formulas depict the goodness of fit and are mapped to color saturation.
 %%
 A saturated color shows a strong correlation.
 %%
 (b) Since $Z$ assumes all features $n$ as given by the formula, it yields $n$ heatmap visualizations.
 %%
 These represent the slices in our cube visualization.
 %%
 The R$^2$ values of each slice voxel is mapped on opacity in the 3D-view, reducing the occlusion of other values.
 }
  \label{fig:Cube}
\end{figure}
%%
The goal of a overview visualization is to provide a comprehensive view on the data, which is easy to understand.
%% TODO: Cite another paper here!
As described in \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded using color on a mosaic plot.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{Regression Cube} analysis we are interested in the goodness of fit of the resulting model, which allows to infer about the predictive quality of the independent features included in the model.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the R$^2$ value is the metric allowing for this kind of assessment.
%%
\paragraph{2D (Slice) View.}
%%
Since R$^2$ is scaled between [0,1] it also allows for comparison between regression models.
%%
We can apply the same mosaic plot mapping by translating the R$^2$ values to color saturation (Fig.~\ref{fig:Cube} a).
%%
This describes a 2D regression square for dynamic variables $X$ and $Y$ (e.g. $Age \sim X + Y$).
%%

\paragraph{3D (Cube) View.}
%%
Introducing $Z$ creates a 3D \emph{Regression Cube} (Fig.~\ref{fig:Cube} b).
%%
R$^2$ values of each cube entry (\emph{voxel}) is mapped to opacity to reduce the overlap.
%%
The visualization of R$^2$ values derived from different regression cubes (e.g. $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
Therefore, the R$^2$ results of different regression methods are encoded using different colors (e.g. blue for linear regression and red for logistic regression).
%%
This way, the cube can easily be extended using other regression types.
%%
For cubes having one fixed target feature, such as $Cancer \sim X + Y + Z$ no such encodings is required and the $z$ dimension can be compared directly.
%%
\\\\
The goal is to create a overview visualization for a data set, but on the other hand we also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
Different analysis approaches require different starting points using the \emph{Regression Cube}.

@@

@@

The design space of web-technologies is different to standard WIMP-applications--right click menus, modal windows and menu bars are not established UI components in this context.

@@

The \emph{Regression Cube} design focuses on a clean interface, reducing the amount of ui elements as much as possible.

@@

To allow other users even outside the epidemiological application domain access to our tool, we choose to do this using standard ASCII-based CSV files.

>>>>

%% Paper title.
%% TODO: Replace towards with regarding?

@@

\maketitle
\com{Bernhard: Comment on prior occurrences of "Regression Cube" and how we differ}
%%
Epidemiology aims to characterize health and disease conditions in defined populations (\emph{Cohorts}).
%%
Insights about risk factors allow to characterize disease-specific high-risk groups and act as important diagnostic key figures \cite{Fletcher2012}.
%%
They can also be used to give recommendations regarding a healthy lifestyle and provide information about wide-spread diseases.
%%
In the standard workflow, physicians translate observations into hypotheses, which are depicted using epidemiological features and then assessed using regression analyses.

An important epidemiological tool for deriving such features are \emph{Cohort studies}, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}.
%%
To reduce any selection bias, subjects are invited at random and without a focus on a specific disease.
%%
The acquired features range from social and lifestyle factors to prior or current diseases and medications as well as medical parameters, such as blood pressure and also comprises non-radiating medical image data. e.g. magnetic resonance imaging (MRI).
%%
Medical image data may be quantified using user-defined landmarks, describing for example shape, volume or diameter of a structure.
%%

To assess the statistical resilience of a hypotheses using regression analyses rarely involves more than three features due to the required subject count.
%%
Due to missing overview techniques, possibly interesting correlations lie within the data, but are not apparent.
%%
Explorative analyses and overview visualizations of the data set as presented by Klemm et al. \cite{Klemm2014VIS} are not tailored to a specific target variable and mostly highlight correlations between variables, which are known to the domain expert (e.g. correlation between body size and spine shape).
%%
We incorporate the regression analysis, which is familiar to the domain experts into overview visualizations, which can either be used for an hypothesis-free analysis or an analysis towards a specific disease or hypotheses.
%%
This is achieved by providing template regression formulas, which are applied to all potential variable combinations.
%%
Since the notation is familiar to epidemiologists, they can rapidly include their domain knowledge into the analysis process.
%%
Difference views between regression formulas allow to assess the influences of individual variables in the process.

Our contributions are:
\begin{itemize}
	\item An overview visualization technique representing feature interactions using target features.
	\item Visualization techniques, which incorporate overview visualizations of all regression analyses at once as well as details on demand techniques for in-depth investigations of feature relationships.
	\item freely adjustable regression formulas provide a simple, yet powerful way to adjust the regression analysis to specific hypotheses about the data.
	\item analysis of confounding variables by providing comparison views between different formula results
	\item The open and web based approach of the system allows for analysis of any data using the presented method.
\end{itemize}

\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow and requirements.
%%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
Epidemiological research is performed by experts from different academic disciplines, such as physicians, statisticians and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is to derive disease-specific risk factors by assessing epidemiological features using statistical methods.
%%
As described by Thew et al. \cite{Thew2009} the epidemiological workflow is divided into these different steps:
\begin{itemize}
	\item Clinicians make observations in the daily practice which are translated into hypotheses.
	\item Epidemiologists compile a list of variables depicting the hypothesis and include confounding variables.
	\item Statisticians assess the association of the derived features towards the investigated disease.
\end{itemize}
%%
Relative risks can be determined if a statistical resilient association of features towards a condition are extracted.
%%
They indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is an epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS} are used to analyze the data using the classical sequential epidemiological workflow, which uses images largely only to communicate results, rather than providing insight.
%%
An alternative data-driven approach is described as follows.

\paragraph{Data Driven Hypothesis-Generation-based Analysis.}
We already described an Interactive Visual Analysis approach for image-centric cohort study data, which connects to the variable listing step \cite{Klemm2014VIS}.
%%
Their methods aim to derive hypotheses through analysis of the data and observations about previously unknown feature correlations.
%%
Employing \emph{hypotheses generation} requires overview visualizations of feature correlations, which are not supported by standard statistical processors.

\subsection{Epidemiological Data} \label{sec:EpidemiologicalData}
%%
\com{Bernhard: Due to overlaps to VIS'14 paper in this section, this can be shortened. E.g. "More discussion of data types and characteristics can be found in [...]"}
%%
\emph{Cohort studies} are a major epidemiological tool for gathering epidemiological features.
%%
They yield a highly heterogenous and incomplete information space.
%%
Subject data are collected either towards a specific disease or with the widest range possible.
%%
The latter allows the data set to be assessed towards different diseases.
%%
The feature space comprises information about lifestyle, somatometric variables, medical parameters, genetic data as well as medical image data.
These features are derived through different modalities, such as questionnaires or medical examinations or laboratory analyses.
%%
Many features are sparse, such as follow up questions about a medication or treatment of a certain disease.
%%
Other features are exclusive for a sub-group, such as women-specific questions, e.g. number of born children or period status.

@@

\paragraph{Image Data.}
\com{Can probably be cut (reference to Klaus' Paper for more details on this) (see Bernhards comment above)}
%%
Since the Rotterdam study, many modern cohort studies include medical image data.
%%
For ethical reasons, the imaging modalities usually not include ionizing radiation.
%%
The image quality is often inferior to clinical standards, which is a tradeoff between time and cost \cite{Preim2014}.
%%
These data are hard to analyze as they require segmentation highlighting the structures of interest.
%%
This process is prone to inter and intra-observer variability when carried out manually.
%%
Automatic or semi-automatic solutions are a remedy for this problem, but are costly and need to be tailored to the structure of interest \cite{Toennies2015}.
%%
Image-derived variables are either dichotomous variables indicating a medical finding by a radiologist or continuous, describing the shape of segmented structures (e.g. volume or diameter).

\paragraph{Confounding Features}
Features influencing the exposure as well as the outcome of a analysis are called confounders.
%%
The analysis model has to be adjusted by normalizing all included features towards the confounding feature.
%%
\emph{Age} is included as confounder in almost any epidemiological analysis, since most diseases (such as different cancer types) are more likely with increasing age.
%%
It also influences the general body condition and therefore almost all features acquired through cohort studies.
%%
Another important confounder is \emph{Gender}.
%%
Other confounder have to be selected by epidemiologists specific to the investigated condition.

\subsection{Regression Analysis} \label{sec:RegressionAnalysis}
%%
Regression analysis is the most important statistical tool when analyzing epidemiological data.
%%
They are also the foundation of this work.
%%
A regression analysis assesses the influence of one or more (\emph{independent}) features to one target (\emph{dependent}) feature.
%%
The regression model yields a function describing the target feature by weighting the independent features.
%%
Metrics, such as the weightings itself and associated p values and describe the resulting function (the \emph{model}).
%%
R$^2$ values describe the goodness of fit; in other words how well the dependent features describe the target feature.
%%
The value is in the range [0,1], while 1 encodes a perfect fit.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany aims to aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per subject.
%%
A second cohort, \texttt{SHIP-TREND} was introduced in 2012.
%%
Data for both cohorts are examined in a in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the last acquisition (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study ("\emph{method loyality}").
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
Tukey already stated in 1977 that data if too often only analyzed using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
We put each paper in relation to our work, derive ideas and point out differences, starting with the works closest to ours:
%%
\\\\
The work of Piringer et al. \cite{Piringer}, Chan et al. \cite{Chan}, Angelelli et al. \cite{Angelelli} and Zhang et al. \cite{Zhang2014} is closest to ours regarding different aspects.
%% HyperMoVal: Interactive Visual Validation of Regression Models for Real-Time Simulation
Akin to our approach, Piringer et al. \cite{Piringer} propose methods for visualizing regression analysis results and properties for developing car engines.
%%
Their main goal is to pairwise assess the influence of independent features towards the target.
%%
This is achieved by using a plot matrix displaying models as contours towards the target feature.
%%
They also incorporate 3D visualizations for each pairwise combination, but mainly because it is popular with the engineers in their target domains.
%%
Linked views of model deviations allow to select outlier residuals.
%%
Comparison of models is also included by mapping model values on line-style (e.g. solid for $model_a$ and dashed for $model_b$).
%%
This limits the method to comparing a few models at once.
%%
The plot matrix representation however gets complex and harder to understand with increasing number of independent features.
%%
The main difference is their focus on analyzing one complex model in detail, yielding in extensive plots, while we analyze a large amount of simpler models towards \emph{different} features.
%%
They also focus on metric features, while we process also categorical data.
%%
We share the matrix-arrangement of results depicting relationships between independent features in a model.

@@

%% TODO: Check DLPD for correct page numbers
While the presented techniques were well received with epidemiological experts, the explanatory power towards back pain was limited, yielding a analysis based on image-derived features, such as extracting and analyzing curvature, torsion and angle of the lumbar spine \cite{Klemm2015}.
%%
These measures were analyzed using a novel decision tree quality plot towards non-image data.
%%
We concluded that the model quality is not enough to characterize back pain.
\\\\
The difference of the presented approach towards our previous and other herein presented work is twofold:
\begin{enumerate}
	\item We focus on large scale analysis of a vast number of linear and logistic regression-models by assessing their quality-of-fit using descriptive metrics.
	\item The analysis is conducted towards a target feature and incorporates expert knowledge via the regression model definition rather than subdividing the underlying data.
\end{enumerate}
\com{Main Difference to our and other related work}
\section{Regression Cube Analysis of Cohort Study Data}
\com{Add Data Preprocessing Section.}
%%
The basic idea of our \emph{Regression Cube} is to provide an overview visualization of large cohort study data sets towards target variables.
%%
Overview visualizations of feature relationships as presented by Angelelli et al. \cite{Angelelli} are often focused on relationships between the visualized features.
%%
Correlation metrics, such as the \emph{Pearson product-moment correlation coefficient} or \emph{Cram\'{e}r's V} contingency values are incorporated to achieve this goal.
%%
In epidemiology, these relationships are also of interest, but rather w.r.t. their explanatory power towards the target feature.
%%
These target features often indicates the presence of the investigated disease.
%%
As described in Section \ref{sec:RegressionAnalysis}, regression analyses are the statistical tool of choice for analyzing these relationships.
%%
A regression model is based on expert knowledge, there is no unified rule on how to apply them to a given set of features, so they have to be applied with care.
%%
\subsection{Cube Description using Regression Formula Notation}
%%
Expert knowledge is introduced into regression analyses using the regression formulas.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the formula input influences the type of the chosen regression method as well as the \emph{independent} features describing the target.
%%
%In order to calculate all combinations of two to three independent features, we can apply  \emph{Regression Cube} aims to 
%%

Since we want to use the regression analyses with an overview visualization, we are are interested in all possible combinations of (two or more) independent features describing a target.
%%
We achieve this by introducing dynamic variables $X$, $Y$ and $Z$ into the regression notation.
%%
The method then replaces the dynamic variables with all features in the data set.
%%
In a data set with $n$ (e.g. 100) features, the regression formula
\begin{equation}
Cancer \sim X + Y
\end{equation}
would yield $n^2$ (10.000) regression models, describing all possible combinations of two features in the data describing the target $Cancer$.
%%
The major advantage of this notation is that it comes natural to anyone familiar with regression analysis, because it uses the same notation, all operators can be used as before.
%%
This allows for a fast adaptation and in the epidemiological application domain.
%%
With simple adjustments to the formula, different results can be achieved:
%%
\begin{itemize}
	\item $Z \sim X + Y$ calculates all combinations of two features towards all possible target features.
	\item $Cancer \sim X + Y + BodyWeight$ includes the $BodyWeight$ feature into all regression models as feature.
	\item $Cancer \sim X + Y + Z$ calculates all combinations of three features towards the target.
\end{itemize}
%%
The problem with this brute-force approach lies in its complexity.
%%
The number of calculated regression models increases exponentially for each dynamic variable added.
%%
If we assume a data set with 100 features with the formula $Z \sim X + Y$ we calculate 1,000,000 regression models.
%%
When each regression takes about 50~ms calculation, we have to wait roughly 14 hours for the calculation to complete.
%%
Therefore, the computational complexity needs to be reduced.
%%
%\com{Regression formulas are described using standard regression formula notation by introducing variables.}
%%
%\com{Target Group: Physicians/Epidemiologists with basic statistical background to understand regression formulas and Interaction.}

\subsection{Target-Variable-dependent Dimension Reduction}
\com{Uli, Can you please proofread this?}
%%
The vast majority of features in an epidemiological data have no or a very low relation towards the target feature.
%%
Identifying these features und excluding them from the calculation can reduce the number of dimension significantly.
%%
The \emph{Correlation based Feature Selection} (CFS) algorithm is very popular in data mining for this task \cite{CFS}.
%%
It uses information entropy to select the features which have the most explanatory power towards the target feature.
%%
At the same time it tries to reduce the correlation between the independent features.
%%
When for example the \emph{body weight} has a strong explanatory power towards the target, it is likely that \emph{BMI} or \emph{waist circumference} behave similar towards it.
%%
They, however, correlate strongly with each other.
%%
The CFS algorithm would then select the feature which has the largest explanatory power and discards the other dimensions.

@@

The number of features calculated by the CFS algorithm is dependent on the information entropy in the data.
%%
In our epidemiological data we observed a usual number of 10 to 20 features.
%%
For features, such as age or gender, which affect most other features (Recall "Confounder" in Section~\ref{sec:EpidemiologicalData}) the number is significantly larger (about half the features in the whole data set).
%%
The number of features extracted is not directly associated with the overall number, but rather on their information entropy towards each other.
%%
\com{Uli, is this correct this way?}
%%

@@

\subsection{Abstracting Regression Results using R$^2$}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/cube_sketch}
 \caption{
 (a) Overview visualization using a heatmap for the formula $Z \sim X + Y$, where $Z$ assumes feature $Age$.
 %%
 The R$^2$ values extracted from the regression formulas depict the goodness of fit and are mapped to color saturation.
 %%
 A saturated color shows a strong correlation.
 %%
 (b) Since $Z$ assumes all features $n$ as given by the formula, it yields $n$ heatmap visualizations.
 %%
 These represent the slices in our cube visualization.
 %%
 The R$^2$ values of each slice voxel is mapped on opacity in the 3D-view, reducing the occlusion of other values.
 }
  \label{fig:Cube}
\end{figure}
%%
The goal of an overview visualization is to provide a comprehensive view on the data (raw or using descriptive metrics \cite{Bertini}), which is easy to understand.
%% TODO: Cite another paper here!
As described in \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded using color on a mosaic plot.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{Regression Cube} analysis we are interested in the goodness of fit of the resulting model, which allows to infer about the predictive quality of the independent features included in the model.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the R$^2$ value is the metric allowing for this kind of assessment.
%%
\paragraph{2D (Slice) View.}
%%
Since R$^2$ is scaled between [0,1] it also allows for comparison between regression models.
%%
We can apply the same mosaic plot mapping by translating the R$^2$ values to color saturation (Fig.~\ref{fig:Cube} a).
%%
This describes a 2D regression square for dynamic variables $X$ and $Y$ (e.g. $Age \sim X + Y$).
%%

\paragraph{3D (Cube) View.}
%%
Introducing $Z$ creates a 3D \emph{Regression Cube} (Fig.~\ref{fig:Cube} b).
%%
R$^2$ values of each cube entry (\emph{voxel}) is mapped to opacity to reduce the overlap.
%%
The visualization of R$^2$ values derived from different regression cubes (e.g. $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
Therefore, the R$^2$ results of different regression methods are encoded using different colors (e.g. blue for linear regression and red for logistic regression).
%%
This way, the cube can easily be extended using other regression types.
%%
For cubes having one fixed target feature, such as $Cancer \sim X + Y + Z$ no such encodings are required and the $z$ dimension can be compared directly.
%%
\\\\
The goal is to create a overview visualization for a data set, but on the other hand we also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
Different analysis approaches require different starting points using the \emph{Regression Cube}.

@@

%% TODO: Umschreiben

@@

The design space spanned by web-technologies is different to standard WIMP-applications--right click menus, modal windows and menu bars are not established UI components in this context.

@@

The \emph{Regression Cube} design focuses on a clean interface, reducing the amount of user-interface elements as much as possible.

@@

To allow other users even outside the epidemiological application domain to access to our tool, we choose to do this using standard ASCII-based CSV files.
